---
title: "Changes in Influenza Antiviral Prescribing in Scotland throughout the COVID-19 Pandemic"
output: html_document
date: "2025-11-14"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Introduction 
The COVID-19 pandemic and its associated non-pharmaceutical interventions (NPIs) drastically altered the transmission dynamics of respiratory infections worldwide. Influenza incidence dropped to historically low levels in 2020–2021, largely due to social distancing, mask mandates, and travel restrictions. As normal social activity resumed, influenza activity rebounded in subsequent seasons. 
Examining national community prescribing of influenza antivirals (oseltamivir, zanamivir) provides a population-level indicator of influenza burden. 

##Library

```{r}
library(tidyverse)
library(here)
library(stringr)
```

## Loading Data Set

```{r}
load_year <- function(year) {
  months <- sprintf("%02d", 1:12)
  dfs <- vector("list", length(months))
  
  for (i in seq_along(months)) {
    m <- months[i]
    path <- here("data", year, paste0("pitc", year, m, ".csv"))
    df <- read_csv(path, show_col_types = FALSE)

    if (!"DMDCode" %in% names(df)) {
      df <- df %>% mutate(DMDCode = NA_character_)
    } else {
      df <- df %>% mutate(DMDCode = as.character(DMDCode))
    }
    
    dfs[[i]] <- df
  }
  bind_rows(dfs)
}
pitc2019_all <- load_year(2019)
pitc2020_all <- load_year(2020)
pitc2021_all <- load_year(2021)
pitc2022_all <- load_year(2022)
pitc2023_all <- load_year(2023)
pitc2023_all <- pitc2023_all %>% select(-PrescribedType)
pitc2024_all <- load_year(2024)
```

## Methods
Filtering Influzenza-related Drug:
https://bnf.nice.org.uk/treatment-summaries/influenza/#related-drugs
```{r}
flu_name <- regex("oseltamivir|tamiflu|zanamivir|relenza", ignore_case = TRUE)

flu2019 <- pitc2019_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month)

## 2020
flu2020 <- pitc2020_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month) 

## 2021
flu2021 <- pitc2021_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month)

## 2022
flu2022 <- pitc2022_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month)

## 2023
flu2023 <- pitc2023_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month)

## 2024
flu2024 <- pitc2024_all %>%
  mutate(
    year  = PaidDateMonth %/% 100,
    month = PaidDateMonth %% 100
  ) %>%
  filter(
    str_starts(BNFItemCode, "050305") |
      str_detect(BNFItemDescription, flu_name)
  ) %>%
  group_by(year, month)

```


```{r}
pitc_all <- bind_rows(
   flu2019, flu2020, flu2021, flu2022, flu2023, flu2024
) %>% 
  arrange(year, month)

flu_monthly <- pitc_all %>%
  group_by(year, month) %>%
  summarise(items = sum(NumberOfPaidItems, na.rm = TRUE), .groups = "drop") %>%
  arrange(year, month)
```

## Plots

```{r}
ggplot(flu_monthly, aes(x = month, y = items,
                        group = year, colour = factor(year))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = 1:12) +
  facet_wrap(~ year, ncol = 2) +
  labs(
    x = "Month",
    y = "Paid items (national total)",
    title = "Prescirptions of Influenza antivirals in Scotland (2019 - 2023)",
    subtitle = "Oseltamivir/Tamiflu & Zanamivir/Relenza",
    colour = "Year"
  ) +
  theme_light()+
  theme(plot.title = element_text(hjust = 0.5),
plot.subtitle = element_text(hjust = 0.5))
```